NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-01-16 main.tex

NMRlipids Databank is a community-driven database containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. It has been designed to improve data Findability, Accessibility, Interoperability, and Reuse (FAIR)~\cite{wilkinson16} using the NMRlipids project protocol where all the content is openly accessible from the beginning of the project~\cite{botan15}.  The NMRlipids databank contains currently nearly thousand simulation trajectories with the total length approaching half a millisecond. The distribution of available simulations containing a specific lipid in Fig.~\ref{fig:overlay}C roughly resembles the biological abundance of different lipid types with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids depending on organism and organelle~\cite{vanmeer08}.Currently available single component lipid membranes and binary mixtures in the NMRlipids databank are illustrated in Fig.~\ref{fig:overlay}D. While these comprise the majority of available trajectories, mixtures with up to five lipid types are available. Force fields used in simulations cover all the essential parameters commonly used in lipid simulations, see Fig.~\ref{fig:overlay}E, including also united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.

@@

%\subsection{Analyzing the databank}
More flexible analyses that enable wide range of novel applications can be performed by the Python application programming interface (API) available at \url{https://github.com/NMRLipids/Databank}. The API can be used to automatically analyse virtually any property from all simulations in the NMRlipids databank following the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}A. Iteration through the README.yaml files in the {\it databank layer} enables the access to raw simulation data via the links stored in these files. Affiliation of universal molecule and atom names to the simulation specific names in README.yaml and mapping files enable automatic analyses independently on the specific naming convention used in a simulation. Analysis codes and results for membrane area per lipid, C–H bond order parameters, X-ray scattering form factors and thickness are included in the {\it Databank layer}. Any other analyses can be implement by users and the results can be stored to the application layer.

@@

\begin{figure}[tbp]
    \centering
    \includegraphics[width=88mm]{Figures/flowchart2.pdf}
    \caption{Differences in area per lipids of POPC and POPE bilayers between different force fields.  
    }
    \label{fig:POPCPOPEapls}
\end{figure}

\subsection{Selection of best simulation parameters for applications using the NMRlipids databank}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
Understanding of how closely simulations resemble the reality is crucial when assessing the reliability and significance of MD simulation results, as well as when using atomistic simulations in parametrization of coarse grained models. However, universal quality measures for MD simulations are lacking, and a prevailing force field that would perform best in all situations has not been found for membranes. For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. 
%and the complicated landscape of force field quality for membranes hamper the applicability of such simulations, resulting in substantial controversies~\cite{antila22b}. 
%For example, previous quality evaluations of membrane simulations have concluded that 
On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, whileGROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. Furthermore, Fig.~\ref{fig:POPCPOPEapls}B illustrates divergent predictions for the difference in packing between POPC and POPE membranes from different force fields. To streamline the selection of best simulations for specific applications, and to guide the development of atomistic and coarse grained parameters, quantitative quality measures for lipid bilayer simulations are defined in the NMRlipids databank.
%that can be used to rapidly find the best available parameters .

The quality for each simulation in the NMRlipids databank with the corresponding experimental data available is currently done using C--H bond order parameters from NMRexperiments and form factors from X-ray scattering experiments. The C--H bond order parameters are sensitive to conformational ensembles of individual lipid molecules~\cite{ollila16}, while correlating well also with the membrane thickness and lateral packing as shown in Fig.~\ref{fig:quality}F. The X-ray scattering form factors are related to membrane dimensions \textit{via} the electron density profile. In particular, locations of minima in these form factors are correlated with the membrane lateral packing and thickness as shown in Figs.~\ref{fig:quality}F,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}.
For quality measures, probabilities for each order parameter to locate within experimental error and distance of the first form factor minimum from the experimental value are calculated from a simulation.
%For detailed definitions of quality measures, see the supplementary information.

@@

In summary, the automatic quality evaluation of the NMRlipids databank enables the rapid exploration and selection of the best simulation parameters for specific applications without extensive and tedious manual force field evaluation. This possibility will promote more reliable simulation results for wide range of applications, support force field development, and streamline the parametrization of coarse grained force fields against the best atomistic MD simulations. The automatically extracted quantitative quality measures will be particularly useful to guide automatic parametrization procedures.

>>>>

NMRlipids Databank is a community-driven database containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. It has been designed to improve data Findability, Accessibility, Interoperability, and Reuse (FAIR)~\cite{wilkinson16} using the NMRlipids project protocol where all the content is openly accessible from the beginning of the project~\cite{botan15}.  Currently the NMRlipids databank contains nearly thousand simulation trajectories with the total length approaching half a millisecond. The distribution of available simulations containing a specific lipid in Fig.~\ref{fig:overlay}C roughly resembles the biological abundance of different lipid types with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids depending on organism and organelle~\cite{vanmeer08}. Currently available single component lipid membranes and binary mixtures in the NMRlipids databank are illustrated in Fig.~\ref{fig:overlay}D. While these comprise the majority of available trajectories, mixtures with up to five lipid types are available. Force fields used in simulations cover all the essential parameters commonly used in lipid simulations, see Fig.~\ref{fig:overlay}E, including also united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.

@@

%\subsection{Analyzing the databank}
More flexible analyses that enable wide range of novel applications can be performed by the Python application programming interface (API) available at \url{https://github.com/NMRLipids/Databank}. The API can be used to automatically analyse virtually any property from all simulations in the NMRlipids databank following the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}A. Iteration through the README.yaml files in the {\it databank layer} enables the access to raw simulation data via the links stored in these files. Affiliation of universal molecule and atom names to the simulation specific names in README.yaml and mapping files enable automatic analyses independently on the specific naming convention used in a simulation. Analysis codes and results for membrane area per lipid, C–H bond order parameters, X-ray scattering form factors and thickness are included in the {\it Databank layer}. Further analysis codes can be implemented by users and the results can be stored to the {\it application layer} in the same format that is used in the {\it Databank layer}. The results from analyses can be then programmatically accessed and furhter analyzed using the NMRlipids databank API following the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}B.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=88mm]{Figures/flowchart2.pdf}
    \caption{ A) Flowchart for analysing properties from all MD simulations in the NMRlipids databank.
    B) Flowchart for accessing the results analyzed from the NMRlipids databank.
    C) Area per lipids of POPC and POPE lipid bilayers predicted by different force fields.  
    }
    \label{fig:POPCPOPEapls}
\end{figure}

\subsection{Selection of best simulation parameters for applications using the NMRlipids databank}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
Understanding of how closely simulations resemble the reality is crucial when assessing the reliability and significance of MD simulation results, as well as when using atomistic simulations in parametrization of coarse grained models. However, universal quality measures for MD simulations are lacking, and a prevailing force field that would perform best in all situations has not been found for membranes. For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. 
%and the complicated landscape of force field quality for membranes hamper the applicability of such simulations, resulting in substantial controversies~\cite{antila22b}. 
%For example, previous quality evaluations of membrane simulations have concluded that 
On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. Furthermore, Fig.~\ref{fig:POPCPOPEapls}B illustrates divergent predictions for the difference in packing between POPC and POPE membranes from different force fields. To streamline the selection of best simulations for specific applications, and to guide the development of atomistic and coarse grained parameters, quantitative quality measures for lipid bilayer simulations are defined in the NMRlipids databank. Together with the API, these will enable rapid and automatic quality evaluation of simulations in the NMRlipids databank, thereby promoting the selection of best simulations for specific applications without laborious force field evaluation, and automatic parametrization procedures for atomistic and coarse grained simulations.
%that can be used to rapidly find the best available parameters .

The quality for each simulation in the NMRlipids databank with the corresponding experimental data available is currently done using C--H bond order parameters from NMR experiments and form factors from X-ray scattering experiments. The C--H bond order parameters are sensitive to conformational ensembles of individual lipid molecules~\cite{ollila16}, while correlating well also with the membrane thickness and lateral packing as shown in Fig.~\ref{fig:quality}F. The X-ray scattering form factors are related to membrane dimensions \textit{via} the electron density profile. In particular, locations of minima in these form factors are correlated with the membrane lateral packing and thickness as shown in Figs.~\ref{fig:quality}F,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}.
For quantitative quality measures, probabilities for each order parameter to locate within experimental error, and the distance of the first form factor minimum from the experimental value are calculated from a simulation.
%For detailed definitions of quality measures, see the supplementary information.

@@

%In summary, the automatic quality evaluation of the NMRlipids databank enables the rapid exploration and selection of the best simulation parameters for specific applications without extensive and tedious manual force field evaluation. This possibility will promote more reliable simulation results for wide range of applications, support force field development, and streamline the parametrization of coarse grained force fields against the best atomistic MD simulations. The automatically extracted quantitative quality measures will be particularly useful to guide automatic parametrization procedures.
